---
title: 'Simulation study'
bibliography: /Users/asingh/Dropbox/Manuscript/diablo/mybib.bib
thanks: "Replication files are available at https://github.com/singha53/diablo"
output:
  html_document:
    theme: united
    highlight: tango
    keep_md: yes
biblio-style: /Users/asingh/Dropbox/Manuscript/diablo/genome-biology.csl
---

```{r setup, message=FALSE, warning=FALSE, tidy=TRUE, echo=FALSE}
knitr::opts_chunk$set(fig.pos = 'p', message=FALSE, warning=FALSE, tidy=TRUE) # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)

WhereAmI <- "~/Dropbox/Manuscript/diablo/analyses/simulation_study/"

## load libraries
library(cowplot)
library(mixOmics)
library(mvtnorm)
library(NMF)

## source functions
source(paste0(WhereAmI, "src/simulation_study_functions.R"))
```

# Simulation study

## Objectives
* how does DIABLO compares with existing integrative classification schemes (Concatenation and Ensemble)
* effect of the design (connection between datasets) on the types of variables selected (correlated vs. discriminatory) and error rate

\pagebreak

## Generate different types of variables and apply diablo to each type separately
* 3 datasets (effective sample size = 100; group1=100 observations, group2=100 observations)
* each dataset has four types of variables; lets explore them now
  + 30 variables that contribute to correlated & discriminatory components
  + 30 variables that contribute to correlated & nondiscriminatory components
  + 100 variables that contribute to uncorrelated & discriminatory components
  + 100 variables that contribute to uncorrelated & nondiscriminatory components

### Correlation structure for each set of simulated variables

#### corDis

```{r}
J=3; fc=3; n=50
Y <- factor(rep(c("group 1", "group 2"), each = n))
## Generate variates that contribute to the full design
rho = 0.8
sigma = matrix(rho, J, J); diag(sigma) = 1
## Variate for Group 1
bComp1 <- as.data.frame(rmvnorm(n, rep(-fc/2, J), sigma))
## Variate for Group 2
bComp2 <- as.data.frame(rmvnorm(n, rep(fc/2, J), sigma))
bComp.full <- rbind(bComp1, bComp2) 
bComp_full_irrelevant <- as.data.frame(rmvnorm(2*n, rep(0, J), sigma))
colnames(bComp.full) <- colnames(bComp_full_irrelevant) <- paste("Dataset", 1:J, sep="_")

pairPlot(bComp.full, group = Y)
```

#### corNonDis

```{r}
pairPlot(bComp_full_irrelevant, group = Y)
```

#### unCorDis

```{r}
## Generate variates that contribute to the null design
rho = 0
sigma = matrix(rho, J, J); diag(sigma) = 1
 # Variate for Group 1
bComp1 <- as.data.frame(rmvnorm(n, rep(-fc/2, J), sigma))
## Variate for Group 2
bComp2 <- as.data.frame(rmvnorm(n, rep(fc/2, J), sigma))
bComp.null <- rbind(bComp1, bComp2)  #+ matrix(rnorm(2*n*J, mean = 0, sd = noise), nc = J)
bComp_null_irrelevant <- as.data.frame(rmvnorm(2*n, rep(0, J), sigma))
colnames(bComp.null) <- colnames(bComp_null_irrelevant) <- paste("Dataset", 1:J, sep="_")

pairPlot(bComp.null, group = Y)
```

#### unCorNonDis

```{r}
pairPlot(bComp_null_irrelevant, group = Y)
```


## Simulation: vary noise and fold-change and compare with other schemes (concatenation/ensembles)
* concatenation_splsda, ensemble_splsda, diabloFull, diabloNull

```{r}
J <- 3
fcSeq <- c(0.5, 1, 2, 3)
noiseSeq <- c(0.2, 0.3, 0.4, 0.5)
fc.noise.grid <- expand.grid(fcSeq, noiseSeq)
colnames(fc.noise.grid) <- c("FC", "Noise")
varsList <- errList <- list()
nperms <- 5
J=3; n=100; p_relevant=30; p_irrelevant=100
folds=10; nrepeat = 1; ncomp = 1; cpus = 1

varsList_repeat <- errList_repeat <- list()
for(z in 1 : nperms){
  for(h in 1 : nrow(fc.noise.grid)){
  fc <- fc.noise.grid[h, "FC"]
  noise <- fc.noise.grid[h, "Noise"]
  
  ## Generate data given fc and noise
  genDats <- simData(fc=fc, noise=noise, J, n, p_relevant, p_irrelevant)
  data <- genDats$data
  Y <- genDats$Y
  
  ## Apply DIABLO full design
  design <- matrix(1, nrow = J, ncol = J)
  diag(design) <- 0
  keepX = list(rep(p_relevant, 1), rep(p_relevant, 1), rep(p_relevant, 1))
  rownames(design) <- colnames(design) <- names(keepX) <- names(data)
  result.full = block.splsda(X = data, Y = Y, ncomp = 1, keepX = keepX, design = design)
  diabloFullPanels <- selectVar(result.full)[1:J] %>% lapply(., function(i){ i[[1]] }) %>% unlist(.)
  cv.full <- perf(result.full, validation = "Mfold", folds = folds, nrepeat = nrepeat, progressBar = FALSE)
  
  ## Apply DIABLO null design
  design <- matrix(0, nrow = J, ncol = J)
  rownames(design) <- colnames(design) <- names(data)
  result.null = block.splsda(X = data, Y = Y, ncomp = 1, keepX = keepX, design = design)
  diabloNullPanels <- selectVar(result.null)[1:J] %>% lapply(., function(i){ i[[1]] }) %>% unlist(.)
  cv.null <- perf(result.null, validation = "Mfold", folds = folds, nrepeat = nrepeat, cpus = cpus, progressBar = FALSE)
  
  ## Concatenation
  concat <- splsda(X = do.call(cbind, data), Y = Y, keepX = rep(J*p_relevant, 1), ncomp = 1)
  concatPanels <- selectVar(concat)$name
  cv.concat <- perf(concat, validation = "Mfold", folds = folds, nrepeat = nrepeat, progressBar = FALSE)
  
  ## Ensemble
  ensemblePanels <- lapply(data, function(i){
    result <- splsda(X = i, Y = Y, keepX = rep(p_relevant, 1), ncomp = ncomp)
    selectVar(result)$name
  }) %>% unlist()
  cv.ensemble <- perf.ensemble.splsda(data, Y, keepX = rep(p_relevant, 1), ncomp = ncomp, nrepeat = nrepeat, M = folds)
  
  ## Proportion of selected variables
  panels <- list(DIABLO_Full=diabloFullPanels, DIABLO_Null=diabloNullPanels,
                 Concatenation=concatPanels, Ensemble=ensemblePanels)
  
  vars <- lapply(panels, function(i){
    as.character(sapply(strsplit(i, "\\."), function(i) i[1]))})
  dats <- lapply(panels, function(i){
    as.character(sapply(strsplit(diabloFullPanels, "_"), function(i) paste(i[-1], collapse = "_")))})
  varDat <- cbind(unlist(dats), unlist(vars)) %>% 
    as.data.frame() %>% 
    mutate(dataset=V1, varType = V2, Classifier = rep(names(dats), sapply(dats, length))) %>% 
    dplyr::select(dataset:Classifier)
  varDat$FC <- paste("FC", fc, sep = "_")
  varDat$Noise <- paste("Noise", noise, sep = "_")
  varsList[[h]] <- varDat %>% group_by(varType, Classifier, FC, Noise) %>% 
  summarise(n = n()) %>% 
  complete(varType, nesting(FC, Noise, Classifier),
    fill = list(n = 0))
  
  ## Error rates
  err <- data.frame(error = c(cv.full$WeightedVote.error.rate$max.dist["Overall.BER", ],
                            cv.null$WeightedVote.error.rate$max.dist["Overall.BER", ], 
                            cv.concat$error.rate$BER[, "max.dist"], cv.ensemble$Mean)) %>% 
         mutate(Classifier = c("DIABLO_Full", "DIABLO_Null", "Concatenation", "Ensemble"))
  err$FC <- paste("FC", fc, sep = "_")
  err$Noise <- paste("Noise", noise, sep = "_")
  errList[[h]] <- err
  }
  varsList_repeat[[z]] <- do.call(rbind, varsList)
  errList_repeat[[z]] <- do.call(rbind, errList)
}

```

## Error rates

```{r, fig.height=10, fig.width=10}
errorRates <- do.call(rbind, errList_repeat)  %>% 
    mutate(Classifier = factor(Classifier, 
    levels = c("DIABLO_Full", "DIABLO_Null", "Concatenation", "Ensemble"))) %>% 
  group_by(Classifier, FC, Noise) %>% 
  summarise(Mean = mean(error), SD = sd(error)) %>% 
  ggplot(aes(x = Classifier, y = Mean)) +
  geom_bar(stat = "identity") + 
  geom_errorbar(aes(ymin = Mean-SD, ymax = Mean+SD)) +
  facet_grid(FC ~ Noise) +
  customTheme(sizeStripFont = 25, xAngle = -45, hjust = 0, vjust = 1, 
              xSize = 15, ySize = 15, xAxisSize = 15, yAxisSize = 15) +
  ylab("Mean±SD of error rate \n 10-fold cross-validation over 50 simulations") +
  xlab("Integrative Classifiers") +
  geom_hline(yintercept = 0.50, linetype = "dashed")
errorRates
```

### Selected variables

```{r, fig.height=10, fig.width=10}
selectedVars <- do.call(rbind, varsList_repeat) %>% ungroup %>% 
    mutate(Classifier = factor(Classifier, 
    levels = c("DIABLO_Full", "DIABLO_Null", "Concatenation", "Ensemble"))) %>% 
  group_by(Classifier, FC, Noise, varType) %>% 
  summarise(Mean = mean(n), SD = sd(n)) %>% 
  ggplot(aes(x = Classifier, y = Mean, color = varType, fill = varType)) +
  geom_bar(stat = "identity") + 
  #geom_errorbar(aes(ymin = Mean, ymax = Mean+SD), position="dodge") +
  facet_grid(FC ~ Noise) +
  customTheme(sizeStripFont = 25, xAngle = -45, hjust = 0, vjust = 1, 
              xSize = 15, ySize = 15, xAxisSize = 15, yAxisSize = 15) +
  ylab("Average number of selected variables \n across 3 datasets over 50 simulations") +
  geom_hline(yintercept = 90, linetype = "dashed") + xlab("Integrative Classifiers") +
  scale_y_continuous(breaks=c(30,60,90))
selectedVars

pdf(paste0(WhereAmI, "results/simulationResults.pdf"), width = 20, height = 10)
plot_grid(errorRates, selectedVars, labels = c("A", "B"), label_size = 30)
dev.off()
```
